A Simplified Plackett-Luce Likelihood Function for Rank Orders

### 2025-11-24 inst/exponential_BT.Rmd
more work

```{r}
x_try
length(support)
support <- support-max(support)
support <- pmax(support, -10)
support  <- matrix(support,length(r_try),length(x_try))
contour(r_try, x_try, support, xlab="r", ylab="x", levels = -c(1,(1:5)*2), lwd=c(1,3,1,1,1,1))
abline(v=1:9, lwd=0.3, col='gray')
abline(v = r_true)
abline(v = 1:9, col = 'gray', lty = 3, lwd = 0.3)
filled.contour(r_try, x_try, support, xlab = "r", ylab = "x")
```

@@

```{r showlike, cache=FALSE}
options(digits=5)
like
alllike <- Reduce("*",like)
alllike 
alllike/max(alllike)
S <- log(alllike)
S <- S-max(S)
S <- pmax(S,-15)
contour(S,levels = -(0:6)*2, lwd=c(1,4,1,1,1,1,1))
filled.contour(S)
```

>>>>

```{r plottoycontour}
x_try
length(support)
support <- support-max(support)
support <- pmax(support, -10)
support  <- matrix(support,length(r_try),length(x_try))
wanted <- x_try >= 0.5
worker <- function(...){
contour(r_try, x_try[wanted], support[,wanted], xlab="r", ylab="x", levels = -c(1,(1:5)*2), lwd=c(1,3,1,1,1,1))
points(r_true,x_true,pch=16,col='red',cex=2)
jj <- support[,wanted]
maxplace <- which(jj==max(jj), arr.ind=TRUE)
points(r_try[maxplace[1]], x_try[wanted][maxplace[2]], pch=16, col='blue', cex=2)
}
worker()
pdf(file="toyexample.pdf")
worker()
dev.off()

filled.contour(r_try, x_try, support, xlab = "r", ylab = "x")
```

@@

```{r showlike, cache=FALSE}
options(digits=5)
like
alllike <- Reduce("*",like)
alllike 
alllike/max(alllike)
S <- log(alllike)
S <- S-max(S)
S <- pmax(S,-15)
contour(a_try, x_try, S, levels = -(0:6)*1, lwd=c(1,4,1,1,1,1,1))

worker <- function(...){
    wanted <- a_try < 0.5
    contour(a_try[wanted], x_try, S[wanted,], levels = -(0:6)*2, lwd=c(1,4,1,1,1,1,1))
    jj <- S[wanted,]
    m <- which(jj == max(jj), arr.ind=TRUE)
    points(a_try[wanted][m[1]], x_try[m[2]], pch=16, cex=2, col='red')
}
worker()
pdf(file="formula1.pdf")
worker()
dev.off()
filled.contour(S)
```

### 2025-11-24 inst/hankin-exponential_BT.tex
more work

\begin{equation}
  \mathcal{L}(x,r)=
  \begin{cases}
    \frac{x^1}{x^1+x^2+x^3}\left(\frac{x^2}{x^1+x^3} + \frac{x^3}{x^1+x^2}\right)\qquad r=1\\
    \frac{x^2}{x^1+x^2+x^3}\left(\frac{x^1}{x^2+x^3} + \frac{x^3}{x^1+x^2}\right)\qquad r=2\\
    \frac{x^3}{x^1+x^2+x^3}\left(\frac{x^1}{x^2+x^3} + \frac{x^2}{x^1+x^3}\right)\qquad r=3
\end{cases}
\end{equation}

@@

\begin{eqnarray}
\mathcal{L}(x;r=1)
\quad &=& \quad\hphantom{+}
            \plackett{2}{3}{1}{4}\nonumber\\
&{}&\quad + \plackett{2}{4}{1}{3}\nonumber\\
&{}&\quad + \plackett{3}{2}{1}{4}\nonumber\\
&{}&\quad + \plackett{3}{4}{1}{2}\nonumber\\
&{}&\quad + \plackett{4}{2}{1}{3}\nonumber\\
&{}&\quad + \plackett{4}{3}{1}{2}
\end{eqnarray}

@@

@@

Here a simplification of the Plackett--Luce likelihood functions for
order statistics was proposed, some simple properties obtained, and
inference problems from three disciplines given.  The results seem to
be difficult to obtain any other way.  The assumptions furnish a
reasonable, intuitive, and computationally tractable likelihood
function with wide applicability.  Further work might include the
analysis of more than one focal competitor.

>>>>

\begin{equation}\label{aftersim}
  \mathcal{L}(x,r)=
  \begin{cases}
    \frac{x^1}{x^1+x^2+x^3}\left(\frac{x^2}{x^1+x^3} + \frac{x^3}{x^1+x^2}\right)\qquad r=1\\
    \frac{x^2}{x^1+x^2+x^3}\left(\frac{x^1}{x^2+x^3} + \frac{x^3}{x^1+x^2}\right)\qquad r=2\\
    \frac{x^3}{x^1+x^2+x^3}\left(\frac{x^1}{x^2+x^3} + \frac{x^2}{x^1+x^3}\right)\qquad r=3
\end{cases}
\end{equation}

@@

\begin{eqnarray}
\mathcal{L}(x;r=1)
\quad &=& \quad\hphantom{+}
            \plackett{2}{3}{1}{4}\nonumber\\
&{}&\quad + \plackett{2}{4}{1}{3}\nonumber\\
&{}&\quad + \plackett{3}{2}{1}{4}\nonumber\\
&{}&\quad + \plackett{3}{4}{1}{2}\nonumber\\
&{}&\quad + \plackett{4}{2}{1}{3}\nonumber\\
&{}&\quad + \plackett{4}{3}{1}{2}\label{requalfour}
\end{eqnarray}

@@

\section{The case of unknown $r$}

Although figure~\ref{1234third} shows a perfectly respectable
log-likelihood function, in practice we often do not know the value of
$r$, and need a joint likelihood function for $(r,x)$.  It is
convenient to normalize $r$ and use $s=r/n$ for consistency between
different values of $n$, and to treat $s$ as a continuous variable.
The only reasonable way to calulate such a likelihood function appears
to be numerical; fortunately this is straightforward.

We leverage the {\tt hyper2} R package~\cite{hankin2017_rnw} which
includes extensive numerical functionality to simulate Plackett-Luce
situations, and in this case function {\tt rrace()} simulates a race
between competitors of known Plackett-Luce strengths.  To obtain a
likelihood function for a specific value of $(x,s)$ at a particular
observed rank $r$, we run repeated {\em in-silico} races between
competitors of Plackett-Luce strengths $x^1,x^2,\ldots x^n$, and
observe the frequency with which competitor $\lceil ns\rceil$ places
$r$.  For example, consider $x=0.1, r=3$.  We might simulate this as
follows:

\label{rsession}
\begin{verbatim}
n <- 9  # number of competitors
s <- 4  # observation: focal competitor places fourth

v <- seq_len(n)
rrank(15, p=0.1^v, as.character(v))
#> A ranktable:
#>       c1 c2 c3 c4 c5 c6 c7 c8 c9
#>  [1,] 2  1  4  3  5  6  7  8  9 
#>  [2,] 1  2  3  5  4  6  7  9  8 
#>  [3,] 1  2  3  4  5  6  7  8  9 
#>  [4,] 1  2  3  4  5  6  7  8  9 
#>  [5,] 1  2  4  3  5  6  7  9  8 
#>  [6,] 1  2  3  4  5  6  7  8  9 
#>  [7,] 1  3  2  4  5  6  7  8  9 
#>  [8,] 1  2  3  4  5  7  6  8  9 
#>  [9,] 3  1  2  5  4  9  6  7  8 
#> [10,] 1  2  3  4  5  6  7  8  9 
#> [11,] 1  2  3  4  5  6  7  8  9 
#> [12,] 1  2  3  5  4  6  7  9  8 
#> [13,] 1  3  2  4  5  6  7  8  9
#> [14,] 1  2  3  4  5  7  6  8  9 
#> [15,] 1  2  5  4  4  6  8  7  9 
\end{verbatim}

Above we see 15 independent Plackett-Luce races; for example the first
one indicates that competitor 2 came first, 1 came second, 4 came
third, and so on.  Looking at the fourth column [headed \verb+c4+,
  ``came fourth''], we see that

\begin{equation}
  \mathcal{L}(0.1,3)\propto 1,\qquad
  \mathcal{L}(0.1,4)\propto 11,\qquad
  \mathcal{L}(0.1,5)\propto 3
\end{equation}

The maximum likelihood estimate for $s$ would therefore be $4/9\simeq
0.44$, conditional on $x=0.1$.  To estimate $x$ we would need repeated
observations, and here we note that small values of $x$ lead to
underdispersed results.

To validate the suggested likelihood method, we set $x=0.85$ and
$r=3$, that is, competitor $i$ has strength $\propto 0.85^i,
1\leqslant i\leqslant 9$; the focal competitor has the third strongest
Plackett-Luce strength of $0.85^3\simeq 0.61$.  The results of running
{\tt rrace()} 100 times with $n=9$ are given in
table~\ref{results100insilico}.

\begin{table}
  \centering
  \begin{tabular}{llllllllll}
    place & 1  & 2  &  3 &  4 &  5 &  6 &  7 & 8 &  9  \\
    count & 14 & 13 & 18 & 12 & 13 &  9 & 11 & 8 &  2
  \end{tabular}
  \caption{Results of repeated \label{results100insilico}
    Plackett-Luce racing with $x=0.85, r=3$; thus the focal competitor
    is the third strongest.  We see that the focal competitor placed
    first 14 times, second 13 times, and so on}
\end{table}

Now, we can run repeated simulations like those on page~\pageref{rsession}, but
with $n=10^4$ trials; as support we use $x$ running from 0.1 to 1 and
$r=1,2,\ldots 9$: figure~\ref{toyexample}.  We see reasonable
agreement between the true values of $x$ and $r$ and the estimates.
Note the odd shape of the contours and the non-independence of the
parameters.

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{toyexample}  % toyexample.pdf produced by exponential_BT.Rmd
\caption{Contours of log-likelihood for $x$ and $r$.  Red point shows
  true values of parameters, blue dot shows maximum likelihood\label{toyexample}
  estimate}
\end{centering}
\end{figure}

\section{Formula 1}

Formula 1 racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  In a typical season, $\sim 20$
competitors race at $\sim 20$ venues and here we consider the 2023
season, using Lewis Hamilton as the focal competitor.  His results are
shown in table~\ref{lewishamilton}; Qatar and the US are disregarded
as he did not finish.

\begin{table}
  \centering
  \begin{tabular}{lllllllll}
venue & BHR & SAU & AUS & AZE & MIA & MON & ESP & CAN\\
place & 5   & 5   & 2   & 6   & 6   & 4   & 2   & 3  \\ \\
venue & AUT & GBR & HUN & BEL & NED & ITA & SIN & JPN\\ 
place & 8   & 3   & 4   & 4   & 6   & 6   & 3   & 5  \\ \\
venue & MXC & SAP & LVG & ABU \\
place & 2   & 8   & 7   & 9
\end{tabular}
  \caption{Lewis Hamilton's results for the 2023 season. \label{lewishamilton}
    We see that he finished fifth in Bahrain, fifth in Saudi Arabia, second in Austria, and so on}
\end{table}

Again using numerical simulation, amounting to repeated evaluation of
generalised version of equations \ref{aftersim} and \ref{requalfour}
(this time with $b=10^4$ trials per point), we obtain
figure~\ref{formula1}.  From this, we see that $r$ is less than about
0.25 and $x$ is between about 0.4 and 0.8.  Recall that, in this
model, $x=0.5$ implies that the Plackett-Luce strength of each
competitor is about twice that of the next.

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{formula1}  % formula1.pdf produced by exponential_BT.Rmd
\caption{Contours of log-likelihood functions \label{formula1}.  Red circle shows
  maximum likelihood estimate}
\end{centering}
\end{figure}

@@

Here a simplification of the Plackett-Luce likelihood functions for
order statistics was proposed, some simple properties obtained, and
inference problems from three disciplines given.  The results seem to
be difficult to obtain any other way.  The assumptions furnish a
reasonable, intuitive, and computationally tractable likelihood
function with wide applicability.  Further work might include the
analysis of more than one focal competitor.
